---
title: "A1_cyclists_and_rain_in_Auckand"
author: "ray_wang"
date: "2020.05.18"
output:
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.height=3)
```

#Use data on the number of people counted cycling in Auckland, and the amount of rain to answer the following qUestions. 

##1. If you try to convert the cycle count data to tidy format (which you don't have to do for this assignment, because it's hard), what is one obstacle you will encounter?

First, I read the data into R.

```{r}
#import packages
options(warn = -1)
library(dplyr)
library(tidyverse)
library(lubridate)
library(ggplot2)
library(s20x)
#read the data into R
cycle2016.df = read.csv(file= "dailyakldcyclecountdata2016_updated.csv", header = TRUE)
cycle2017.df = read.csv(file = "dailyakldcyclecountdata2017_1.csv", header = TRUE)
cycle2018.df = read.csv(file = "dailyakldcyclecountdata2018.csv", header = TRUE)
#we can see the brief information about these three dataframes
#head(cycle2018.df)
#head(cycle2017.df)
#head(cycle2016.df)
ncol(cycle2018.df)
ncol(cycle2017.df)
ncol(cycle2016.df)

rain2018.df = read.table(file = "rain2018.txt", header = TRUE, sep = ",")
rain16.17.df = read.table(file = "rain2016-17.txt", header = TRUE, sep = ",")
ncol(rain2018.df)
ncol(rain16.17.df)

```

I find the columns of cyclist data are different in 2016,2017,2018.  
So I can't bind these three years data into one directly. 
There are differnt names for same counters in these threee years data.
If I try to convert the cycle count data to tidy format, the obstacle is the difficulty in counter names change.

##2. Compute the total number of cyclists counted for each day, and a suitable summary of the rainfall for each day, in the same data frame.
Before I start to manipulate the data, I must transform the NA into 0.  


```{r}
cycle2018.df[is.na(cycle2018.df)]<-0
cycle2017.df[is.na(cycle2017.df)]<-0
cycle2016.df[is.na(cycle2016.df)]<-0
cycle2018.df$Total.number.Cyclists <- rowSums(cycle2018.df[,2:ncol(cycle2018.df)])
cycle2017.df$Total.number.Cyclists <- rowSums(cycle2017.df[,2:ncol(cycle2017.df)])
cycle2016.df$Total.number.Cyclists <- rowSums(cycle2016.df[,2:ncol(cycle2016.df)])
```

Then get the sum of cyclists each day.
```{r}
cycle2018.df$Total.number.Cyclists <- rowSums(cycle2018.df[,2:ncol(cycle2018.df)])
cycle2017.df$Total.number.Cyclists <- rowSums(cycle2017.df[,2:ncol(cycle2017.df)])
cycle2016.df$Total.number.Cyclists <- rowSums(cycle2016.df[,2:ncol(cycle2016.df)])
```

It is necessary to tidy the data. 
We need seprate the date into four variables: day of week, day of month, month and year.
And it helps to have numeric values stored as numeric values rather than as strings.We can mutate them.
We pick the useful columns from the data which includes the date and the sum of cyclist each day.
After we get three years data tidy, we can bind them together.

``` {r}
cycle_18 <-cycle2018.df %>%
  separate(col=Date, into=c("dow", "day", "month", "year")) %>%
  mutate(dayno=as.numeric(day), yearno=as.numeric(year),
         wday=factor(dow,levels=c("Mon","Tue","Wed","Thu","Fri","Sat","Sun")))
#tail(cycle_18)
#there is a 2019.0101 data which is useless in this assignment, we should delete this row. 
cycle_18 = cycle_18[-366,]
#pick the useful columns from the data which includes the date and the sum of cyclist each day.
cycle_18 = cycle_18[,-c(5:(ncol(cycle_18)-4))]
#tail(cycle_18)

#the procedure is same as above
cycle_17 <-cycle2017.df %>%
  separate(col=Date, into=c("dow", "day", "month", "year")) %>%
  mutate(dayno=as.numeric(day), yearno=as.numeric(year),
         wday=factor(dow,levels=c("Mon","Tue","Wed","Thu","Fri","Sat","Sun")))
cycle_17 = cycle_17[,-c(5:(ncol(cycle_17)-4))]
#tail(cycle_17)

#the procedure is same as above
cycle_16 <-cycle2016.df %>%
  separate(col=Date, into=c("dow", "day", "month", "year")) %>%
  mutate(dayno=as.numeric(day), yearno=as.numeric(year),
         wday=factor(dow,levels=c("Mon","Tue","Wed","Thu","Fri","Sat","Sun")))
cycle_16 = cycle_16[,-c(5:(ncol(cycle_16)-4))]
#tail(cycle_16)

cycle = rbind(cycle_16,cycle_17,cycle_18)
summary(cycle)
head(cycle)
tail(cycle)

```

Then tidy the rain data. Aggregate the rain amount each hour to get the rain amount each day.
Similarly, we We need seprate the date into three variables: day of month, month and year.
And it helps to have numeric values stored as numeric values rather than as strings.We can mutate them.
After we get the rain data tidy, we can bind the rain data andy cyclist data together.
```{r}
#head(rain2018.df)
#head(rain16.17.df)
summary(rain2018.df)
summary(rain16.17.df)
rain_tmp = rbind(rain16.17.df, rain2018.df)
# Aggregate the hours rain amount to get the rain amount each day.
rain_sum<-aggregate(Amount.mm. ~  Date.NZST.,data=rain_tmp,sum)

#tidy the rain data 
rain <- rain_sum %>%
  select(Date=Date.NZST., Rainfall=Amount.mm.) %>%
  separate(Date, into=c("year","month","day"),sep=c(4,6)) %>%
  mutate(dayno=as.numeric(day))
  
#there is a 2019.0101 data which is useless in this assignment, we should delete this row. 
rain = rain[-1097,]
tail(rain)
Rainfall =  rain$Rainfall
monthno = rain$month
#confirm that the number of rows is equal in these two dataframe 
nrow(rain)
nrow(cycle)
#After we get the rain data tidy, we can bind the rain data andy cyclist data together.
cycle_rain <- cbind(cycle , Rainfall, monthno)%>%
  mutate(monthno =as.numeric(monthno))%>%
  #transfrom month to season
  mutate(season = quarter(monthno))


```

We achieve the goal: compute the total number of cyclists counted for each day, and a suitable summary of the rainfall for each day, in the same data frame.
```{r}
summary(cycle_rain)
head(cycle_rain)
```


##3. Draw suitable graphs to display how the number of cyclists varies over time, over season, over day of the week, and with rain

Use qpplot() to see the trend over time, over day of the week, over season and with rain. 
```{r}
ggplot(cycle_rain, aes(Rainfall,Total.number.Cyclists)) + geom_point()
qplot(Rainfall, Total.number.Cyclists,data=cycle_rain, col=wday, 
      main="The relation between Rainfall and the number of cyclists\n (colors indicate the day of week)")
trendscatter(Total.number.Cyclists ~ Rainfall,main = "The relation between Rainfall and the number of cyclists" ,data=cycle_rain)
qplot(wday,Total.number.Cyclists,data=cycle_rain, col=wday, geom= "boxplot",
      main="The relation between the day of week and the number of cyclists\n (colors indicate Rainfall)")
qplot(month,Total.number.Cyclists,data=cycle_rain, col=month, geom= "boxplot",
      main="The relation between month and the number of cyclists\n (colors indicate Rainfall)")
qplot(factor(season),Total.number.Cyclists,data=cycle_rain, col=season, geom= "boxplot",
      main="The relation between season and the number of cyclists \n(colors indicate Rainfall)")
qplot(year,Total.number.Cyclists,data=cycle_rain, col=Rainfall, geom= "boxplot", 
      main="The relation between year and the number of cyclists\n (colors indicate Rainfall)")

```

It is obvious that the number of cyclists decrease when the rain amount increase.
And people like go to cycle at Tuesday most while they don't like go to cycle at Saturday.
What's more, people like go to cycle in December most while they don't like go to cycle in April, July and June.
In other words, they perfer go to cycle in season 4 (month 10-12) while don't like go to cycle in season 3 (7-9) and season 2 (3-6).
In long term, the number of cyclists increase from 2016 to 2018. 

##4. Fit a regression model to predict the number of cyclists from year, season, day of the week, and rain.

Finally, we can fit a model to predict the number of cyclists from year, season, day of the week, and rain.
Accroding to the anaylsis of graphs above, I choose total numberof cyclists each day as the dependent variable and the amount of rain, the day of weeks and season as independent variables.
By the way, the variable month is useful too, but it has the similar impact as the variable season, so I didn't choose month as one of independent variables.

```{r}
model <- cycle_rain %>%
  lm(Total.number.Cyclists~Rainfall + factor(year) + factor(wday)+factor(season),data=.)
coef(summary(model))
summary(model)
```

Almost very variables' p-value is good which means variables in this model are significant. 
The model is acceptable.
The model support the analysis of graphs above that the number of cyclists decrease when the rain amount increase and people prefer go to cycle at Tuesday and in season 4 most while they don't like go to cycle at Saturday, Sunday and in season 3 and season 2.

##5. Based on your graphs and model, does rain have a big impact on the number of people cycling in Auckland?

Yes, I think rain have a big impact on the number of people cycling in Auckland.
First, the p-value of the variable 'Rainfall' is very close to zero which proves its significance.
What's more, the cofficient of variable 'Rainfall' is -229.6 which means the number of cyclists decrease about 230 when the rain amount increase 1 mm in Auckland.
In summary,  rain have a big impact on the number of people cycling in Auckland.



