---
title: "A2_predicting_taxi_tips"
author: "ray_wang"
date: '2020.05'
output:
  html_document:
    df_print: paged
---

## Using the data from week 2 of January 2016 (week2.csv), construct a model that predicts the amount of a tip.  Evaluate the mean squared error of this model on the data from week 4 of January 2016 (week4.csv). 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


```

```{r}
options(warn = -1)
library(dplyr)
library(tidyverse)
library(lubridate)
library(leaflet)
library(leaps)
library(s20x)

```

# data clean

Read the csv data into R.

```{R}
week2 = read.csv(file= "week2.csv", header = TRUE)
week4 = read.csv(file= "week4.csv", header = TRUE)
nrow(week2)
tail(week2)
summary(week2)
```

After reading the description of this taxi data,I find there are some columns that are clearly uninformative. 

I decide to delete extra, mta_tax, improvement_surcharge and because only credit card tips are included, I need to select the data which payment type == 1
For justification, we can give a command to see there is no tip record in other payment_type.

```{R}
week2[which(week2$payment_type != 1& week2$Tip_amount != 0 ),]
```

There are some illegal negative value. Let's do the data clean. 

* Filter rows with illegal values and recode time sensibly. 
* Recode location to pick up airport trips including leave airport and go to airport.
* Recode pickup and dropoff location into categories, evening, morning, afternoon, night.

```{R}
isAirport <- function(x,y){
  longt <- x;
  lat <- y;
  airport_longt = -73.7785;
  airport_lat =  40.6415;
  return( abs(x-airport_longt)<=0.035&abs(y-airport_lat)<=0.025)
}
taxi <- subset(week2, select = -c(extra, mta_tax, improvement_surcharge))%>%
  filter(payment_type == '1', tip_amount >= 0, fare_amount >= 0 ,
         trip_distance >= 0,trip_distance< 2000,tolls_amount>= 0, RatecodeID <= 6)%>%
  separate(col= tpep_dropoff_datetime, into=c("dropoff_day", "dropoff_time"), sep = c('T') )%>%
  separate(col= dropoff_time, into=c("dropoff_time", "rest"), sep = 'Z' )%>%
  unite("dropoff_datetime", dropoff_day, dropoff_time,sep = ' ')%>%
  separate(col= tpep_pickup_datetime, into=c("pickup_day", "pickup_time"), sep = c('T') )%>%
  separate(col= pickup_time, into=c("pickup_time", "rest2"), sep = 'Z' )%>%
  unite("pickup_datetime", pickup_day, pickup_time,sep = ' ')%>%
  select(-c(rest, rest2,payment_type))%>%
  mutate(
    dow = wday(pickup_datetime,label=TRUE,abbr=TRUE, week_start = 1, locale = Sys.setlocale("LC_TIME", "American")),                           
    hour_trip_start = hour(pickup_datetime),                                   
    trip_duration = as.numeric(difftime(dropoff_datetime,pickup_datetime,units="mins")),
    gotoAirport = isAirport(dropoff_longitude,dropoff_latitude),
    leaveAirport =isAirport(pickup_longitude,pickup_latitude)
  )

taxi$WhatTime <- with(taxi,  ifelse(hour_trip_start >= 5 & hour_trip_start<=11, "morning",
                                     ifelse(hour_trip_start>11 & hour_trip_start<=16, "afternoon",
                                            ifelse(hour_trip_start>16 & hour_trip_start<= 19, "evening","night"))))
```

## Draw plots

We can draw the barplot fair quickly, but we can't draw the plot which including many points, like qpplot(). 

```{R}
taxi %>%
  ggplot(aes(x=factor(hour_trip_start))) + geom_bar()
taxi %>%
  ggplot(aes(x=factor(WhatTime))) + geom_bar()
taxi %>%
  ggplot(aes(x=factor(dow))) + geom_bar()
taxi %>% ggplot(aes(trip_duration)) + geom_histogram()
taxi %>% filter(trip_duration < 120 & trip_duration >0) %>% ggplot(aes(trip_duration)) + geom_histogram()


```

I want to see what these trips looks like in the map and draw some point plots.
But the data is too big for my PC, so I have to take a random subsample from the origin data. 

```{R}
sm_taxi <- taxi %>% sample_n(1e4)
min_lat <- 40.577
max_lat <- 40.918
min_long <- -74.15
max_long <- -73.700
sm_taxi <- sm_taxi %>% filter(pickup_longitude > min_long, pickup_longitude < max_long, pickup_latitude > min_lat, pickup_latitude < max_lat)
airport_taxi <-sm_taxi%>%filter(leaveAirport == TRUE)
#draw map in R
leaflet(sm_taxi) %>% 
  addTiles() %>% 
  addCircleMarkers(~pickup_longitude,~pickup_latitude, radius=2,stroke = FALSE, opacity=1, fillOpacity =1)
leaflet(airport_taxi) %>% 
  addTiles() %>% 
  addCircleMarkers(~pickup_longitude,~pickup_latitude, radius=2,stroke = FALSE, opacity=1, fillOpacity =1)
```

From the map, we find there are some airport trips. 
We can infer that the airport trip could be a factor in our model. So we mutate two new columns: leaveAirport and gotoAirport.

Draw some plots to to find potential relationships between variables to include eg time of day/week/distance etc.

```{R}
qplot(trip_distance, tip_amount, col=dow, data = sm_taxi,
               main="The relation between trip distance and tip amount \n (colors indicate the day of week)")
ggplot(sm_taxi, aes(trip_distance, tip_amount)) + geom_hex() 
trendscatter(tip_amount~ trip_distance,main = "The relation between trip_distance and tip amount" ,data = sm_taxi)


qplot(fare_amount, tip_amount, data = sm_taxi,
      main="The relation between fare_amount and tip amount")
ggplot(sm_taxi, aes(fare_amount, tip_amount)) + geom_hex() 
trendscatter(tip_amount~ fare_amount,main = "The relation between fare_amount and tip amount" ,data = sm_taxi)

qplot(trip_duration, tip_amount, col=dow, data = sm_taxi,
      main="The relation between trip duration and tip amount \n (colors indicate the day of week)")


qplot(store_and_fwd_flag, tip_amount, data = sm_taxi,
      main="The relation between store_and_fwd_flag and tip amount ")
qplot(store_and_fwd_flag, tip_amount,data = sm_taxi, geom= "boxplot" , col = store_and_fwd_flag)+ ggtitle("tip_amount VS store_and_fwd_flag ")

qplot(gotoAirport, tip_amount,data = sm_taxi, geom= "boxplot", col = gotoAirport)+
ggtitle("tip_amount VS gotoairport")

qplot(leaveAirport, tip_amount,data = sm_taxi, geom= "boxplot", col = leaveAirport)+
ggtitle("tip_amount VS leaveAirport")

qplot(factor(passenger_count), tip_amount, data = sm_taxi,
      main="The relation between passenger_count and tip amount \n (colors indicate the day of week)")
qplot(factor(passenger_count), tip_amount,data = sm_taxi, geom= "boxplot")+
ggtitle("tip_amount VSpassenger_count")

qplot(dow, tip_amount, data = sm_taxi,geom= "boxplot", col = dow,
      main="The relation between dow and tip amount ")

qplot(WhatTime, tip_amount, data = sm_taxi, geom= "boxplot", col = WhatTime,
      main="The relation between Time and tip amount  ")




```

We find trip_distance, fare_amount, and airport trips has big effect on tip_amount.
The day of week and what time in a day has slight effect on tip_amount.

## Model fit 

Before fit the model, we need clean the train data again and prepare the test data week4. 
The test data also need to be clean because we must have the same columns as the train data wk2 to ensure the prediction works.

```{R}
taxi_lm <- taxi %>%filter(trip_duration >=0)%>%
  select(-c(hour_trip_start, total_amount, dropoff_datetime, dropoff_longitude,
                          dropoff_latitude, pickup_datetime, pickup_latitude, pickup_longitude, VendorID))%>%
  mutate(
    dow = factor(dow),  
    RatecodeID = factor(RatecodeID),
    store_and_fwd_flag = factor(store_and_fwd_flag),
    gotoAirport = factor(gotoAirport),
    leaveAirport = factor(leaveAirport),
    WhatTime = factor( WhatTime)
  )
summary(taxi_lm)

taxi_test <- subset(week4, select = -c(extra, mta_tax, improvement_surcharge))%>%
  filter(payment_type == '1')%>%
  separate(col= tpep_dropoff_datetime, into=c("dropoff_day", "dropoff_time"), sep = c('T') )%>%
  separate(col= dropoff_time, into=c("dropoff_time", "rest"), sep = 'Z' )%>%
  unite("dropoff_datetime", dropoff_day, dropoff_time,sep = ' ')%>%
  separate(col= tpep_pickup_datetime, into=c("pickup_day", "pickup_time"), sep = c('T') )%>%
  separate(col= pickup_time, into=c("pickup_time", "rest2"), sep = 'Z' )%>%
  unite("pickup_datetime", pickup_day, pickup_time,sep = ' ')%>%
  select(-c(rest, rest2,payment_type))%>%
  mutate(
    dow = wday(pickup_datetime,label=TRUE,abbr=TRUE, week_start = 1, locale = Sys.setlocale("LC_TIME", "American")),                           
    hour_trip_start = hour(pickup_datetime),                                   
    trip_duration = as.numeric(difftime(dropoff_datetime,pickup_datetime,units="mins")),
    gotoAirport = isAirport(dropoff_longitude,dropoff_latitude),
    leaveAirport =isAirport(pickup_longitude,pickup_latitude)
  ) 
taxi_test$WhatTime <- with(taxi_test,  ifelse(hour_trip_start >= 5 & hour_trip_start<=11, "morning",
                                              ifelse(hour_trip_start>11 & hour_trip_start<=16, "afternoon",
                                                                                                          ifelse(hour_trip_start>16 & hour_trip_start<= 19, "evening","night"))))

taxi_test <- taxi_test%>%select(-c(hour_trip_start, total_amount, dropoff_datetime, dropoff_longitude,
 dropoff_latitude, pickup_datetime, pickup_latitude, pickup_longitude, VendorID))%>%
  filter(tip_amount >= 0, trip_distance >= 0,trip_distance< 2000,fare_amount >= 0 ,tolls_amount>= 0, trip_duration >=0,RatecodeID <= 6)%>%
  mutate(
    dow = factor(dow),  
    RatecodeID = factor(RatecodeID),
    store_and_fwd_flag = factor(store_and_fwd_flag),
    gotoAirport = factor(gotoAirport),
    leaveAirport = factor(leaveAirport),
    WhatTime = factor( WhatTime)
  )

summary(taxi_test)

```

### Calculate the MSPE and do model selection

Then, fit the model. 
Here I use backwards selection methods to do the model selection. And use R-squared and BIC as the criteria.  
First we should fit a full model. And we should delete *dow* from our model according to plots above.
```{R}
library(car)
taxi.lm = lm(tip_amount ~ ., data =taxi_lm)
summary(taxi.lm)
taxi.pred <- predict(taxi.lm, taxi_test, interval = "prediction", level = 0.95)
#MSPE
mean((taxi_test$tip_amount-taxi.pred[,1])^2)
vif(taxi.lm)

```



*passenger_count* has poor p value. I decide to delete *passenger_count*.

*fare_amount* and *trip_distance* have multicolinearity. I decide to delete *fare_amount*.

The MSPE of our first model isn't very good.

```{r}
taxi.lm_2 = lm(tip_amount ~ .-fare_amount-passenger_count-dow, data =taxi_lm)
summary(taxi.lm_2)
vif(taxi.lm_2)
taxi.pred_2 <- predict(taxi.lm_2, taxi_test, interval = "prediction", level = 0.95)
#MSPE
mean((taxi_test$tip_amount-taxi.pred_2[,1])^2)
```

Now, all variables have good p value and the result of VIF check is good.
But the MSPE is still not good. I need do more model selection. 
Here I use leaps package to do the following steps.

```{r}
leaps <- regsubsets(tip_amount ~ .-fare_amount-passenger_count,data =taxi_lm, nbest = 4 )
reg = summary(leaps)
reg
plot(leaps, scale = "adjr2")
plot(leaps, scale = "bic")

summary(leaps)$which[which.max(reg$bic),]
summary(leaps)$which[which.max(reg$adjr2),]

```

We pick the model has max adjusted R-squared as the final model.
That is: 

```{r}
taxi.lm_final = lm(tip_amount ~ RatecodeID+tolls_amount+trip_distance+leaveAirport + WhatTime, data =taxi_lm)
summary(taxi.lm_final)
vif(taxi.lm_final)
taxi.pred_final <- predict(taxi.lm_final, taxi_test, interval = "prediction", level = 0.95)
#MSPE
mean((taxi_test$tip_amount-taxi.pred_final[,1])^2)
```

The MSPE is about 4.8, which is a little high, that means the error of every trip tip amount is about 2 in week4 data. 

## Interpret model and accuracy
* The p-value of every variable in the model is good enough. 
* Generally, variables in the model are sensible. And the number of variables is 5, which is easy to understand and do analysis.
* The face validity is OK, but not very good. The prediction is plausiable. For example, the 1 unit trip_distance increase, about 0.47 tip amount increase. The 1 unit tolls_amount increase, about 0.22 tip amount increase. Passengers with ratecodeID 6 (group ride) prefer pay about 8.4 more dollars tip. But the airport factors show negative effect on tip amount, which is not very sensible.  
* The R-squared is 0.5086,which means the model can explain about 51% data in week2 data. 

However, after model selection, the accuracy doesn't improve much. 
The MSPE is about 4.8, that means the error of every trip tip amount is about 2 in week4 data. 

It is possible that linear model without interaction variable can't fit the data well.
We can add interaction term into the model and do model selection again. 


