---
title: "A2_predicting_taxi_tips"
author: "ray_wang"
date: '2020.05'
output:
  html_document:
    df_print: paged
---

## Using the data from week 2 of January 2016 (week2.csv), construct a model that predicts the amount of a tip.  Evaluate the mean squared error of this model on the data from week 4 of January 2016 (week4.csv). 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


```

```{r}
options(warn = -1)
library(dplyr)
library(tidyverse)
library(lubridate)
library(leaflet)
library(leaps)
library(s20x)

```

# data clean

Read the csv data into R.

```{R}
week2 = read.csv(file= "week2.csv", header = TRUE)
week4 = read.csv(file= "week4.csv", header = TRUE)
nrow(week2)
tail(week2)
summary(week2)
```

After reading the description of this taxi data,I find there are some columns that are clearly uninformative. 

I decide to delete extra, mta_tax, improvement_surcharge and because only credit card tips are included, I need to select the data which payment type == 1
For justification, we can give a command to see if there is tip record in other payment_type.
There are a few of tip records in other payment_type but most of them are illegal.
So I decide to just pick the data which *payment_type* == 1.

```{R}
week2[which(week2$payment_type != 1& week2$tip_amount != 0 ),"tip_amount"]
```

There are some illegal negative value. Let's do the data clean. 

* Filter rows with illegal values and recode time sensibly. 
* Recode location to pick up airport trips including leave airport and go to airport.
* Recode pickup and dropoff location into categories, evening, morning, afternoon, night.
* Recode date into weekdays and weekend.

```{R}
isAirport <- function(x,y){
  longt <- x;
  lat <- y;
  airport_longt = -73.7785;
  airport_lat =  40.6415;
  return( abs(x-airport_longt)<=0.025&abs(y-airport_lat)<=0.020)
}

isWeekend <- function(dow){
  return( dow == "Sat" | dow == "Sun")
}
taxi <- subset(week2, select = -c(extra, mta_tax, improvement_surcharge))%>%
  filter(payment_type == '1', tip_amount >= 0, fare_amount >= 0 ,tip_amount < 1000,
         trip_distance >= 0,trip_distance< 1000,tolls_amount>= 0, RatecodeID <= 6)%>%
  separate(col= tpep_dropoff_datetime, into=c("dropoff_day", "dropoff_time"), sep = c('T') )%>%
  separate(col= dropoff_time, into=c("dropoff_time", "rest"), sep = 'Z' )%>%
  unite("dropoff_datetime", dropoff_day, dropoff_time,sep = ' ')%>%
  separate(col= tpep_pickup_datetime, into=c("pickup_day", "pickup_time"), sep = c('T') )%>%
  separate(col= pickup_time, into=c("pickup_time", "rest2"), sep = 'Z' )%>%
  unite("pickup_datetime", pickup_day, pickup_time,sep = ' ')%>%
  select(-c(rest, rest2,payment_type))%>%
  mutate(
    dow = wday(pickup_datetime,label=TRUE,abbr=TRUE, week_start = 1, locale = Sys.setlocale("LC_TIME", "American")),                           
    hour_trip_start = hour(pickup_datetime),                                   
    trip_duration = as.numeric(difftime(dropoff_datetime,pickup_datetime,units="mins")),
    gotoAirport = isAirport(dropoff_longitude,dropoff_latitude),
    leaveAirport =isAirport(pickup_longitude,pickup_latitude),
    isWeekend = isWeekend(dow)
  )

taxi$WhatTime <- with(taxi,  ifelse(hour_trip_start >= 5 & hour_trip_start<=11, "early morning",
                                     ifelse(hour_trip_start>11 & hour_trip_start<=16, "afternoon",
                                            ifelse(hour_trip_start>16 & hour_trip_start<= 19, "evening","night"))))
```

## Draw plots

We can draw the barplot fair quickly, but we can't draw the plot which including many points, like qpplot(). 

```{R}
taxi %>%
  ggplot(aes(x=factor(hour_trip_start))) + geom_bar()
taxi %>%
  ggplot(aes(x=factor(WhatTime))) + geom_bar()
taxi %>%
  ggplot(aes(x=factor(dow))) + geom_bar()
taxi%>%
  ggplot(aes(x=factor(isWeekend))) + geom_bar()
taxi %>% ggplot(aes(trip_duration)) + geom_histogram()
taxi %>% filter(trip_duration < 120 & trip_duration >0) %>% ggplot(aes(trip_duration)) + geom_histogram()


```

I want to see what these trips looks like in the map and draw some point plots.
But the data is too big for my PC, so I have to take a random subsample from the origin data. 

```{R}
sm_taxi <- taxi %>% sample_n(1e4)
min_lat <- 40.577
max_lat <- 40.918
min_long <- -74.15
max_long <- -73.700
sm_taxi <- sm_taxi %>% filter(pickup_longitude > min_long, pickup_longitude < max_long, pickup_latitude > min_lat, pickup_latitude < max_lat)
airport_taxi <-sm_taxi%>%filter(leaveAirport == TRUE)
#draw map in R
leaflet(sm_taxi) %>% 
  addTiles() %>% 
  addCircleMarkers(~pickup_longitude,~pickup_latitude, radius=2,stroke = FALSE, opacity=1, fillOpacity =1)
leaflet(airport_taxi) %>% 
  addTiles() %>% 
  addCircleMarkers(~pickup_longitude,~pickup_latitude, radius=2,stroke = FALSE, opacity=1, fillOpacity =1)
```

From the map, we find there are some airport trips. 
We can infer that the airport trip could be a factor in our model. So we mutate two new columns: leaveAirport and gotoAirport.

Draw some plots to to find potential relationships between variables to include eg time of day/week/distance etc.

```{R}
sm_taxi <- taxi %>% sample_n(1e4)

qplot(trip_distance, tip_amount, col=dow, data = sm_taxi,
               main="The relation between trip distance and tip amount \n (colors indicate the day of week)")
ggplot(sm_taxi, aes(trip_distance, tip_amount)) + geom_hex() 
trendscatter(tip_amount~ trip_distance,main = "The relation between trip_distance and tip amount" ,data = sm_taxi)

qplot(fare_amount, tip_amount, data = sm_taxi,
      main="The relation between fare_amount and tip amount")
ggplot(sm_taxi, aes(fare_amount, tip_amount)) + geom_hex() 
trendscatter(tip_amount~ fare_amount,main = "The relation between fare_amount and tip amount" ,data = sm_taxi)

qplot(trip_duration, tip_amount, col=dow, data = sm_taxi,
      main="The relation between trip duration and tip amount \n (colors indicate the day of week)")


qplot(store_and_fwd_flag, tip_amount, data = sm_taxi,
      main="The relation between store_and_fwd_flag and tip amount ")
qplot(store_and_fwd_flag, tip_amount,data = sm_taxi, geom= "boxplot" , col = store_and_fwd_flag)+ ggtitle("tip_amount VS store_and_fwd_flag ")

qplot(gotoAirport, tip_amount,data = sm_taxi, geom= "boxplot", col = gotoAirport)+
ggtitle("tip_amount VS gotoairport")

qplot(leaveAirport, tip_amount,data = sm_taxi, geom= "boxplot", col = leaveAirport)+
ggtitle("tip_amount VS leaveAirport")

qplot(factor(passenger_count), tip_amount, data = sm_taxi, col = passenger_count,
      main="The relation between passenger_count and tip amount \n ")
qplot(factor(passenger_count), tip_amount,data = sm_taxi, col = passenger_count, geom= "boxplot")+
ggtitle("tip_amount VS passenger_count")

qplot(factor(isWeekend), tip_amount, data = sm_taxi, col = isWeekend,
      main="The relation between isWeekend and tip amount \n ")
qplot(factor(isWeekend), tip_amount,data = sm_taxi, geom= "boxplot")+
ggtitle("tip_amount VS isWeekend")

qplot(dow, tip_amount, data = sm_taxi,geom= "boxplot", col = dow,
      main="The relation between dow and tip amount ")

qplot(isWeekend, tip_amount, data = sm_taxi,geom= "boxplot", col = isWeekend,
      main="The relation between dow and tip amount ")

qplot(WhatTime, tip_amount, data = sm_taxi, geom= "boxplot", col = WhatTime,
      main="The relation between Time and tip amount  ")


qplot(factor(RatecodeID), tip_amount, col= RatecodeID, data = sm_taxi, geom= "boxplot",
               main="The relation between RatecodeID and tip amount \n ")


```

We find *trip_distance*, *fare_amount*, *RatecodeID* and *airport trips* have big effect on *tip_amount*.
The day of week, is weekend or not and what time in a day has slight effect on *tip_amount*.

## Model fit 

Before fit the model, we need clean the train data again and prepare the test data week4. 
The test data also need to be clean because we must have the same columns as the train data wk2 to ensure the prediction works.

```{R}
taxi_lm <- taxi %>%filter(trip_duration >=0)%>%
  select(-c(hour_trip_start, total_amount, dropoff_datetime, dropoff_longitude,
                          dropoff_latitude, pickup_datetime, pickup_latitude, pickup_longitude, VendorID))%>%
  mutate(
    dow = factor(dow),  
    RatecodeID = factor(RatecodeID),
    store_and_fwd_flag = factor(store_and_fwd_flag),
    gotoAirport = factor(gotoAirport),
    leaveAirport = factor(leaveAirport),
    WhatTime = factor( WhatTime),
    isWeekend = factor(isWeekend)
  )
summary(taxi_lm)

taxi_test <- subset(week4, select = -c(extra, mta_tax, improvement_surcharge))%>%
  filter(payment_type == '1')%>%
  separate(col= tpep_dropoff_datetime, into=c("dropoff_day", "dropoff_time"), sep = c('T') )%>%
  separate(col= dropoff_time, into=c("dropoff_time", "rest"), sep = 'Z' )%>%
  unite("dropoff_datetime", dropoff_day, dropoff_time,sep = ' ')%>%
  separate(col= tpep_pickup_datetime, into=c("pickup_day", "pickup_time"), sep = c('T') )%>%
  separate(col= pickup_time, into=c("pickup_time", "rest2"), sep = 'Z' )%>%
  unite("pickup_datetime", pickup_day, pickup_time,sep = ' ')%>%
  select(-c(rest, rest2,payment_type))%>%
  mutate(
    dow = wday(pickup_datetime,label=TRUE,abbr=TRUE, week_start = 1, locale = Sys.setlocale("LC_TIME", "American")),                           
    hour_trip_start = hour(pickup_datetime),                                   
    trip_duration = as.numeric(difftime(dropoff_datetime,pickup_datetime,units="mins")),
    gotoAirport = isAirport(dropoff_longitude,dropoff_latitude),
    leaveAirport =isAirport(pickup_longitude,pickup_latitude),
    isWeekend = isWeekend(dow)
  ) 
taxi_test$WhatTime <- with(taxi_test,  ifelse(hour_trip_start >= 5 & hour_trip_start<=11, "early morning",
                                              ifelse(hour_trip_start>11 & hour_trip_start<=16, "afternoon",
                                                                                                          ifelse(hour_trip_start>16 & hour_trip_start<= 19, "evening","night"))))

taxi_test <- taxi_test%>%select(-c(hour_trip_start, total_amount, dropoff_datetime, dropoff_longitude,
 dropoff_latitude, pickup_datetime, pickup_latitude, pickup_longitude, VendorID))%>%
  filter(tip_amount >= 0, tip_amount <1000,trip_distance >= 0,trip_distance< 1000,fare_amount >= 0 ,tolls_amount>= 0, trip_duration >=0,RatecodeID <= 6)%>%
  mutate(
    dow = factor(dow),  
    RatecodeID = factor(RatecodeID),
    store_and_fwd_flag = factor(store_and_fwd_flag),
    gotoAirport = factor(gotoAirport),
    leaveAirport = factor(leaveAirport),
    WhatTime = factor( WhatTime),
    isWeekend = factor(isWeekend)
  )

summary(taxi_test)

```

### Calculate the MSPE and do model selection

Then, fit the model. 
Here I use backwards selection methods to do the model selection. And use p-value, R-squared, and AIC (Akaike Information Criterion) as the criteria.  
First we should fit a full model. And we should delete *dow* from our model according to plots above.

```{R}
library(car)
taxi.full = lm(tip_amount ~ .-dow, data =taxi_lm)
summary(taxi.full)
vif(taxi.full)
taxi.pred <- predict(taxi.full, taxi_test, interval = "prediction", level = 0.95)
#MSPE
mean((taxi_test$tip_amount-taxi.pred[,1])^2)

```

*passenger_count* has poor p value. I decide to delete *passenger_count*.

According to the VIF check, the problem of colinearity is not significant but do exist.

And thinking of the plots I drew above, *trip_distance* and *fare_amount* have almost same effect on *tip_amount*, I decide to delete *trip_distance* for this variable has the biggest VIF. 

The MSPE of our first model is about 4.5, isn't very good. 

Let's continue the model selection.

```{r}
taxi.lm_2 = lm(tip_amount ~ .-passenger_count-dow-trip_distance, data =taxi_lm)
summary(taxi.lm_2)
vif(taxi.lm_2)
taxi.pred_2 <- predict(taxi.lm_2, taxi_test, interval = "prediction", level = 0.95)
#MSPE
mean((taxi_test$tip_amount-taxi.pred_2[,1])^2)


```

Now, all variables have good p value and the result of VIF check is good.
But the MSPE doesn't improve much. I need do more model selection. 
Here I use regsubsets() from leaps package to do the following steps.

```{r}
leaps <- regsubsets(tip_amount ~ .-passenger_count-dow-trip_distance,data =taxi_lm, nbest = 5 )
reg = summary(leaps)
reg
plot(leaps, scale = "adjr2", main = "R-suqared as criteria")
summary(leaps)$which[which.max(reg$adjr2),]
```

Pick the model has max adjusted R-squared as the next model.
That is: *RatecodeID-3, RatecodeID-5, fare_amount, isWeekend, WhatTime-morning*

```{r}
taxi.lm_3 = lm(tip_amount ~ (RatecodeID==2)+(RatecodeID==3)+(RatecodeID==5)+fare_amount+isWeekend+trip_duration+(WhatTime=="early morning"), data =taxi_lm)
summary(taxi.lm_3)
taxi.pred_3 <- predict(taxi.lm_3, taxi_test, interval = "prediction", level = 0.95)
#MSPE
mean((taxi_test$tip_amount-taxi.pred_3[,1])^2)
```

But the MSPE is not good.

I use *stepAIC()* from MASS package to do model selection again.

```{r}
library(MASS)
taxi.step <- stepAIC(taxi.lm_2, direction = "backward")
summary(taxi.step)
taxi.pred_step <- predict(taxi.step, taxi_test, interval = "prediction", level = 0.95)
#MSPE
mean((taxi_test$tip_amount-taxi.pred_step[,1])^2)
vif(taxi.step)

```
*stepAIC()* and *regsubsets* give us two best models in two criteria.

The MSPE of  the model that *stepAIC()* gives is smaller which has the smallest AIC too. 

I choose this model as the final model.

That is:

```{r}
taxi.lm_final = lm(tip_amount ~  fare_amount + tolls_amount +  gotoAirport+isWeekend+RatecodeID + WhatTime+ isWeekend, data =taxi_lm)
summary(taxi.lm_final)
vif(taxi.lm_final)
taxi.pred_final <- predict(taxi.lm_final, taxi_test, interval = "prediction", level = 0.95)
#MSPE
mean((taxi_test$tip_amount-taxi.pred_final[,1])^2)

```
The MSPE is about 4.5, that means the prediction error of every trip's tip amount is about 2 in week4 data.

## Interpret model and accuracy

* The p-value of every variable in the model is good enough which means all variables are significant.
* Generally, variables in the model are sensible. And the number of variables is 6, which is easy for people to understand.
* The face validity is OK, but not very good. The prediction is plausiable. 
  * The 1 unit fare_amount increase, about 0.16 unit tip amount increase. 
  * The 1 unit tolls_amount increase, about 0.25 unit tip amount increase.
  * Passengers with RatecodeID 6 (group ride) prefer pay about 5.14 dollars tip more.
  * And people who go to airport pay 0.17 dollars tip more. 
  * People who have trip in weekend has pay about 0.1 dollars tip less.
* The R-squared is 0.5423,which means the model can explain about 54% data in week2 data. 

However, after model selection, the accuracy doesn't improve much. 
The MSPE is about 4.5, that means the prediction error of every trip's tip amount is about 2 in week4 data.

It is possible that linear model without interaction variable can't fit the data well.
We can add interaction term into the model and do model selection again. 


